---
title: 'Quality of Ontario Lakes'
subtitle: 'How has the quality of Ontario’s lakes changed overtime? And as of 
recent, how productive are Ontario’s lakes?'
authors: 'Joshua Antonio Crisologo: _____ & Jonathan Manuel: 1010080797'
output: pdf_document
date: "2024-03-30"
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(knitr)
library(lubridate)
```


```{r, include=FALSE}

# loading the data

trans <- read.csv("../data/transparency.csv")
phosphorus <- read.csv("../data/phosphorus.csv")
calcium <- read.csv("../data/calcium.csv")
chloride <- read.csv("../data/chloride.csv")

# data processing 

# Selected date-range is 2015-2022

phosphorus <- phosphorus %>% select(!Data.Collector)

glimpse(calcium)

calcium <- calcium %>% mutate(Date = as.Date(Date..DD.MMM.YY., 
                                             format = "%d-%b-%y")) %>% 
  select(-Date..DD.MMM.YY., -X_id) %>% 
  filter(2015 <= year(Date)) %>% 
  rename(lat = Latitude..DMS., long = Long..DMS., calcium_mg_l = Calcium..mg.L., 
         cal_date=Date) %>% 
  arrange(lat)

chloride <- chloride %>% mutate(Date = as.Date(Date..dd.mm.yyyy., 
                                               format = "%d/%m/%Y")) %>% 
  select(-Date..dd.mm.yyyy., -X_id) %>% 
  filter(2015 <= year(Date)) %>% 
  rename(lat = Latitude..DMS., long = Longitude..DMS., 
         chlroide_mg_l = Chloride..mg.L., chl_date=Date) %>% 
  arrange(lat)

# Dropped Sample 1 & Sample 2 in favor of "Average.Total.Phosphorus..µg.L."
phosphorus <- phosphorus %>% mutate(Date = as.Date(Date..DD.MMM.YY., 
                                                   format = "%d-%b-%y")) %>% 
  select(-Date..DD.MMM.YY., -X_id, -Total.Phosphorus.sample.1..µg.L., 
         -Total.Phosphorus.sample.2..µg.L.) %>% 
  filter(2015 <= year(Date)) %>% 
  rename(lat = Latitude..DMS., long = Long..DMS., 
         avg_phos_ug_l = Average.Total.Phosphorus..µg.L., phos_date=Date, 
         phos_is_outlier = Possible.outlier) %>% 
  arrange(lat)

# Messes everything up; Maybe best to keep it as serperate; Diff lat vals than others

trans <- trans %>% mutate(Date = as.Date(Date..DD.MMM.YY., 
                                         format = "%d-%b-%y")) %>% 
  select(-Date..DD.MMM.YY., -X_id) %>% 
  filter(2015 <= year(Date)) %>% 
  rename(lat = Latitude..DMS., long = Longitude..DMS., 
         secchi_depth_m = Secchi.Depth..m., trans_date=Date, Township=TOWNSHIP) %>% 
  arrange(lat)

# The common variables between data sets
sims = c("lat", "long",  "STN", "Site.ID", 
         "Township","Lake.Name", "Site.Description")

tmp1 <- merge(calcium, chloride, by=sims)
tmp2 <- merge(phosphorus, trans, by=sims)

# The line below breaks RStudio lol

# data <- merge(tmp1, tmp2, by=sims)
            
```


## Introduction 

Sources to add while citing: https://data.ontario.ca/dataset/ontario-lake-partner
& https://files.ontario.ca/moe_mapping/downloads/metadata/opendata/Lake_Partner_Program_metadata_EN2.pdf

This report relies on four datasets sourced from the Ontario Lake Partner 
Program (LPP) via the Ontario Data Catalog. The LPP conducts annual assessments 
of water quality in inland lakes throughout Ontario, with data collected by 
volunteers following standardized provincial protocols. Specifically, the 
datasets cover total phosphorus, calcium, chloride concentration, and water 
clarity for numerous inland lakes, predominantly situated in the Precambrian 
Shield region. Each dataset includes geographical coordinates, site descriptions, 
collection dates, and metrics pertaining to the water quality. The data was last 
validated on January 17, 2024 and is updated yearly. The largest overlapping 
period across all four datasets spans from January 1, 2015, to December 31, 2022.

## Purpose

This report aims to address the research question: 

How has the quality of Ontario’s lakes changed overtime? And as of recent, 
how productive are Ontario’s lakes?

## 

