---
title: 'Quality of Ontario Lakes'
subtitle: 'How has the quality of Ontario’s lakes changed overtime? And as of 
recent, how productive are Ontario’s lakes?'
authors: 'Joshua Antonio Crisologo: _____ & Jonathan Manuel: 1010080797'
output: pdf_document
date: "2024-03-30"
---

```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(knitr)
library(lubridate)
```


```{r, include=FALSE}

# loading the data

transparency <- read.csv("../data/transparency.csv")
phosphorus <- read.csv("../data/phosphorus.csv")
calcium <- read.csv("../data/calcium.csv")
chloride <- read.csv("../data/chloride.csv")

# data processing 

phosphorus <- phosphorus %>% select(!Data.Collector)


```


## Introduction 

Sources to add while citing: https://data.ontario.ca/dataset/ontario-lake-partner
& https://files.ontario.ca/moe_mapping/downloads/metadata/opendata/Lake_Partner_Program_metadata_EN2.pdf

This report relies on four data sets sourced from the Ontario Lake Partner 
Program (LPP) via the Ontario Data Catalog. The LPP conducts annual assessments 
of water quality in inland lakes throughout Ontario, with data collected by 
volunteers following standardized provincial protocols. Specifically, the 
data sets cover total phosphorus, calcium, chloride concentration, and water 
clarity for numerous inland lakes, predominantly situated in the Precambrian 
Shield region. Each data set includes geographical coordinates, site descriptions, 
collection dates, and metrics pertaining to the water quality.

## Purpose

This report aims to address the research question: 

How has the quality of Ontario’s lakes changed overtime? And as of recent, 
how productive are Ontario’s lakes?

## 

